---
title: "Examples for useful data wrangling functions"
author: "Maria Novosolov"
date: "9/21/2021"
output: rmdformats::readthedown
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```


# Classic one's (`dplyr` package)

* `select` - allows you to select or deselect columns

* `filter` - allows you to filter rows by values in some columns

* `mutate` - allows you to change anything, either values within columns or adding new columns with new values



# Some simple but useful

* `clean_names` (from `janitor`) - changes all the column names to easy to deal with R format

* `rename` (from `dplyr`)- lets you rename a column, or many columns

* `na_if` (from `dplyr`) - allows you to replace any pattern with NA (a good way of getting rid of annoying characters that you forgot to remove)

* `drop_na` (from `tidyr`)- allows you to remove all the rows with NA

* `relocate` (from `dplyr`) - allows you to relocate columns in the dataset

* `distinct` - Allows to remove duplicated rows or retain unique values in a column


# Within data manipulations

* `case_when` (from `dplyr`) - allows you to manipulate the data inside the columns based on conditions. Used inside mutate

* `separate` (from `tidyr`) - allows you to separate a column to multiple columns (similar to "text-to-column" option in excel)



# Manipulate data structure

* `pivot_longer` - creates a long format from your dataframe

* `pivot_wider` - creates a wide format from your data

* `group_by` - lets you group the data based on a categorical variable or multiple ones

* `nest` - lets you create a subset of datasets in the same table

* `tally` - allows you to count how many rows you have in a group

* `across` - lets you do operations across multiple columns



# Examples for each of the functions

## `select` 

```{r include=FALSE}
data(iris)
iris<- as_tibble(iris)
```

### Option 1
```{r}
iris %>% 
  select(Petal.Length) 
```


### Option 2
```{r}
iris %>% 
  select(-Petal.Length) 
```


## `filter` 

### Option 1

```{r}
iris %>% 
  filter(Species == "setosa")
```



### Option 2

```{r}
iris %>% 
  filter(Petal.Length > 1.5)
```




## `mutate`

###  Option 1
```{r}
iris %>% 
  mutate(Petal.Length = Petal.Length*100)
```



### Option 2

```{r}
iris %>% 
  mutate(new.Petal.Length = Petal.Length*100)
```





## `janitor::clean_names`

Changes all the column names to easy to deal with R format
```{r}
iris %>% 
  janitor::clean_names()
```



## `dplyr::rename`

### Change one column

```{r}
iris %>% 
  dplyr::rename(plant_species = Species)
```




### Give all columns a new name

```{r}
newnames<- c("X1","X2","X3","X4","X5")
oldnames<- names(iris)
iris %>% 
  dplyr::rename_with(~newnames,oldnames)

```



### Rename based on condition

```{r}
iris %>% 
  rename_with(toupper, starts_with("Petal"))
```



## `dplyr::na_if`

Allows you to replace any pattern with NA (a good way of getting rid of annoying characters that you forgot to remove). 
Used mainly inside mutate

```{r}
iris %>% 
  mutate(species_new = dplyr::na_if(x = Species,y = "setosa"))
```



## `tidyr::drop_na`

```{r}
iris %>% 
  mutate(species_new = dplyr::na_if(x = Species,y = "setosa")) %>% 
  drop_na()
```



## `dplyr::relocate` 

### Option 1

```{r}
iris %>% 
  relocate(Species)
```



### Option 2

```{r}
iris %>% 
  relocate(Species, .after = Sepal.Width)
```


### Option 3

```{r}
iris %>% 
  relocate(Sepal.Length,.after = everything())
```



## `dplyr::distinct` 

Remember: If you remove all duplicates based on column it will ignore that other columns have unique values. `.keep_all = T` is needed to retain the other columns as well

### Option 1
```{r}
iris %>% 
  distinct(Species,.keep_all = T)
```


### Option 2

```{r}
iris %>% 
  select(Species) %>% 
  distinct()
```



## `dplyr::case_when`

```{r}
iris %>% 
  mutate(Species = as.character(Species)) %>% 
  mutate(new_species = case_when(Species=="setosa"~"setosa_who",TRUE~Species))
```



## `tidyr::separate`

```{r}
iris %>% 
  mutate(Species = paste(Species,"something",sep="_"))
```


```{r}
iris %>% 
  mutate(Species = paste(Species,"something",sep="_")) %>% 
  separate(Species,into = c("Speices","extra")) 
```

## `pivot_longer` 

```{r}
iris %>% 
  pivot_longer(cols = -Species,names_to = "leaf_traits",values_to = "value")

```

## `pivot_wider`

```{r}
iris %>% 
  mutate(id = seq(1,nrow(.))) %>% 
  pivot_longer(cols = -c(id,Species),names_to = "leaf_traits",values_to = "value") 
```


```{r}
iris %>% 
  mutate(id = seq(1,nrow(.))) %>% 
  pivot_longer(cols = -c(id,Species),names_to = "leaf_traits",values_to = "value") %>% 
  pivot_wider(names_from = leaf_traits,values_from = value)
```



## `group_by` / `tally` / `across`

Usually coupled with other function like `mutate`, `summarize`, `nest`, or `tally`

```{r}
iris %>% 
  group_by(Species) %>% 
  tally()
```



```{r}
iris %>% 
  group_by(Species) %>% 
  summarize(across(1:4, mean))
```




## `nest`

A powerful tool to create sub datasets within a tibble. Together with `map` you can do multiple operations on it and then unnset it in the end 

```{r}
iris %>% 
  group_by(Species) %>% 
  nest() 
```

